### 2015-04-20 whoop.tex
various updates

We have implemented our approach in \whoop, a static lockset analysis infrastructure that uses state-of-the-art compilation and sequential verification techniques to (i) automatically analyse Linux drivers for potential data races and (ii) exploit the race-related information from the sound static analysis phase to accelerate bug-finding.

@@

Figure~\ref{fig:whoop} depicts the \whoop toolchain. Initially, \whoop accepts a Linux driver written in C, together with a Linux environmental model\footnote{Stub header files that model the Linux kernel APIs.}, which is required to ``close'' the driver and allow it to be subsequently over-approximated and analysed. Both the driver and the environmental model are passed to the translation phase of \whoop (see Section~\ref{whoop:translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}. Next, \whoop performs \emph{static pairwise lockset analysis} to detect all potential data races in the abstract program. The proof sketch behind our static lockset analysis is given in Section~\ref{whoop:proof}. The instrumentation and analysis is described in more detail in Section~\ref{whoop:method}. The summarisation that we perform to achieve scalability is presented in Section~\ref{whoop:summarisation}. After the static lockset analysis phase, \whoop exploits the race-related information to accelerate bug-finding with \corral (see Section~\ref{whoop:bugfinding})

Finally, we discuss optimisations and assumptions in Section~\ref{whoop:optimisations}, and limitations of our technique in Section~\ref{whoop:limitations}.

@@

The main aim of this phase is to take an input Linux driver written in C, together with an environmental model, and output an over-approximated version of the program in Boogie IVL, a simple imperative language with well-defined verification-focused semantics used as the input to a number of cutting-edge verifiers (e.g. Boogie and \corral). To achieve this translation, \whoop uses three LLVM\footnote{http://llvm.org}-based tools: Chauffeur, Clang\footnote{http://clang.llvm.org} and SMACK\footnote{https://github.com/smackers/smack}~\cite{rakamaric2014smack}.

@@

We now formalise our approach to verifying data race freedom in a driver by statically analysing its locksets. Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a driver, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this driver. Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all drivers that we have studied so far in the Linux kernel have only a small number of locks (typically one). Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking. The actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.

@@

The Engine component of \whoop (see Figure~\ref{fig:whoop}) is responsible for parsing, instrumenting and sequentialising the abstract Boogie program. During the initial parsing, \whoop performs a static analysis (on the Boogie IVL code) to identify all lock identifiers and rewrite them to a unique constant Boogie variable, e.g. the following call to \texttt{mutex\_lock()}:

@@

If \whoop cannot infer a lock it will exit with a warning. A limitation of lockset analysis is that it cannot understand external locking mechanisms (\whoop currently only supports Linux kernel mutexes and spinlocks). It is relatively straightforward to enhance \whoop, though, with new locking primitives. \whoop will also exit with a warning if it detects ``improper'' use of locks (e.g. dynamic lock creation). The reason behind this is twofold: first, it is arguably infeasible to detect such locks using static analysis; and second, because we want to advocate the use of good locking practises when developing drivers for the Linux kernel.

@@

\whoop continues by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialisation}, which combines the two arbitrary threads in a single sequential pair. This is achieved by creating a checker function that calls the two instrumented entry points and then performs race checking assertions. This process repeats until all possible pairs of entry points have been sequentialised. To be sound, each time an entry point performs a read access to a shared memory location, \whoop returns a nondeterministic value (using the Boogie keyword \texttt{havoc}). This over-approximates any effects from all the unmodeled threads on the shared state of the pair. Two-thread reduction is not a new idea, it has been used before in GPU kernel verification~\cite{bardsley2014engineering} and in model checking of cache coherence protocols~\cite{mcmillan1999verification}.

@@

Finally the abstract program is send to the Boogie verification engine, which generates verification conditions~\cite{barnett2005weakest} and discharges them to a theorem prover (Boogie supports Z3~\cite{de2008z3} and CVC4~\cite{barrett2011cvc4}). Successful verification implies that the original program is free of data races, while an error denotes a \emph{potential} data race and is reported to the user. To improve usability \whoop has a built-in error reporter that matches Boogie counterexamples to source code. The following is an example of a reported error:

@@

\subsection{Watchdog Summarisation}
\label{whoop:summarisation}

Early on during the development of \whoop, we realised that scalability would be a serious issue. When we tried to apply \whoop on the r8169 RealTek ethernet driver (8000 lines of C code) with all functions fully inlined, we were running out of memory. One reason was that the C to Boogie translation produced hundreds of thousands of Boogie IVL code, which could not fit in memory. Another reason was that some entry points recursively call helper functions. Inherently, recursion does not work with inlining. To tackle these issues we developed \emph{watchdog summarisation}, a novel summarisation technique that allows us to analyse Linux drivers in a scalable fashion, but without destroying precision.

Watchdog summarisation uses the Houdini~\cite{flanagan2001houdini} invariant inference algorithm (available inside Boogie) to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.

The candidate invariants are based on the global variables that \whoop instruments to perform lockset analysis and race checking (see Section~\ref{whoop:method}). The key idea behind watchdog summarisation is that we automatically generate candidate invariants for each memory region in the granularity of watched accesses. This reduces the false positives from reporting errors if only e.g. a field in memory region \texttt{\$M.0} is racing, but the rest of the fields are not. An example of watchdog-based candidate post-conditions is the following:

@@

\whoop is a sound but potentially imprecise static data race analyser.  While
developers do appreciate soundness, it is more important to ensure they are not
overwhelmed by countless reports of false bugs.  Hence, to improve precision, we
enable an external precise bug-finder for concurrent programs to be plugged into
the \whoop infrastructure. The race-related information generated by \whoop can
then be used to accelerate the bug-finder.  Currently, we support
\corral~\cite{lal2012corral}, a state-of-the-art scalable bug-finder used by
Microsoft to analyse Windows device drivers. \corral accepts Boogie IVL and
hence it was easy to integerate into our tool flow. Our technique, though, is
general and capable of accelerating any arbitrary bug-finder for concurrent
programs.

\subsubsection{Bug-Finding with \corral}
\label{bf:corral}

\corral is a symbolic bounded verifier for Boogie IVL that uses SMT solvers
(e.g., Z3 or CVC4) to statically reason about program behaviours. It checks for
violations of provided assertions, and reports a precise counterexample if an
assertion violation is found. \corral performs bounded exploration of a
concurrent program in two steps. First, given a bound on the number of allowed
context-switches, the concurrent program is appropriately \emph{sequentialized},
and the generated sequential version preserves reachable states of the original
concurrent program~\cite{popl2011-eqr,cav2009-lqr,cavLalR08}.  Then, \corral
attempts to prove bounded (in terms of the number of loop iterations and
recursion depth) sequential reachability of a bug in a goal-directed, lazy
fashion to postpone state space explosion when analysing a large program.  It
performs two abstractions to achieve this: (i) variable abstraction, where it
attempts to identify a minimal set of shared variables that have to be precisely
tracked in order to discharge all assertions; and (ii) stratified inlining,
where it attempts to inline procedures on-demand as they are required for
proving program assertions.

@@

By default, \corral instruments a \texttt{yield} statement (which denotes a
potential nondeterministic context-switch) before each \emph{visible operation}
(i.e., synchronisation operations and shared memory accesses) of every thread.
It then leverages sequentialization to explore all possible thread interleavings
up to a pre-defined bound.

Our approach to accelerating \corral is simple and yet effective: we simplify
the sequentialisation that \corral performs by instrumenting \texttt{yield}
statements only at shared memory accesses that \whoop reported as potentially
racy. This effectively acts as a partial order reduction that can significantly
reduce the number of interleavings that \corral needs to explore, and thus
increase performance and scalability.

\ZRComment{Could this be cleaned up a bit? In particular, do we need to talk
about entry point pairs, or could we prove it just for any pair?}
We now show that our partial order reduction is sound: it does not miss any more
bugs than what \corral would miss even without taking advantage of the \whoop
race-related information. Let \corral run an $n$ amount of entry point pairs
with yields in all visible operations. Let this configuration be known as $C$.
Each entry point pair is running in a separate \corral process. \corral will
miss all bugs that are triggered by more than 3 bugs running concurrently and
all bugs that are triggered by investigating a larger than $r$ recursion bound
and larger than $k$ context switch bound. Let \corral then run the same $n$
amount of pairs, but this time with all \texttt{yield} statements in non-racy
shared memory accesses (according to \whoop) removed. Let this configuration be
known as $W+C$. Because these shared memory accesses cannot race, performing a
\texttt{yield} will not introduce any new program behaviours. Thus, \corral will
find the same bugs using the configuration $W+C$, as using the configuration
$C$.

@@

\yieldcoarse instruments \texttt{yield} statement in a binary fashion: if \whoop
finds that a pair of entry points cannot race, then it will only instrument
context-switches in synchronisation points of this pair; else if \whoop finds
that a pair of entry points can race, then it will instrument context-switches
in all visible operations of this pair.

\yieldmr is a finer-grained instrumentation: \whoop instruments context-switches
in all synchronisation points, but only instruments context-switches in memory
regions that can potentially race (regardless if a pair has not been fully
verified as race-free). In our experiments (see Section~\ref{evaluation}),
\yieldmr performs significantly faster than \yieldcoarse.

@@

\subsection{Optimisations and Assumptions}
\label{whoop:optimisations}

A common practise for statically analysing software is to ``close'' the environment by abstracting away the low level implementation details. Any abstract environmental model, though, can and will ultimately result in decreased precision and thus false positives. This is one of the main limitations of statically analysing real life software that use complex external APIs. In our case, we developed a simple environmental model for the Linux kernel that consists of stub header files. Because we only focus on lockset-based properties during the race-checking phase of \whoop, we can get away with abstracting a lot of the underlying functionality. Our model, though, can still result in a lot of false positives.

@@

In our experience, one of the most effective precision optimisations is to enrich \whoop with information regarding kernel imposed locks. The Linux kernel is able to serialise a number of entry point calls, which thus can never run concurrently with each other. An an example, a large number of networking and ethernet entry points are mutually serialised with RTNL, a network-specific kernel lock. \whoop will not create pairs for entry points that cannot run concurrently.

Another optimisation, is to soundly reduce the checked memory locations. If a memory location is accessed only by one entry point in the pair, then it is safe to not check it. This can potentially decrease the amount of generated candidate invariants and speed up the verification process.

Modelling kernel methods is another opportunity to increase precision, but requires expert knowledge about Linux. For example, the \texttt{register\_netdev()} function registers a network driver with the kernel. Before this function is called, the kernel is not able to invoke any of the driver's entry points. Modelling this functionality can reduce false positives when race checking accesses before the driver's registration.

We also optimise the use of function pointers. LLVM and SMACK do not perform context-sensitive translation, which means that a function that accepts a function pointer as an argument, can arbitrarily call any possible method that corresponds to that function pointer regardless of call site. To increase precision, Chauffeur analyses the AST of the original program and attempts to infer context-sensitive function pointer use information. This information is the used by \whoop to increase precision. To be sound, this optimisation is conservative: if no information can be extracted for a function pointer at a call site, then \whoop uses the original over-approximated function.

@@

Environmental modelling is another limitation of \whoop. As aforementioned, enriching \whoop with domain-specific knowledge to increase precision is an ongoing manual effort, but requires expertise in the Linux kernel. We argue that further increasing the precision of our models is orthogonal to the contributions of this paper. Moreover, despite unavoidable imprecisions during the sound static data race analysis, \whoop can use the reported information to significantly speedup precise bug-finders for concurrent programs, as seen in Sections~\ref{whoop:bugfinding} and~\ref{evaluation}.

>>>>

We have implemented our approach in \whoop, a static lockset analysis infrastructure that uses state-of-the-art compilation and sequential verification techniques to (i) automatically analyze Linux drivers for potential data races and (ii) exploit the race-related information from the sound static analysis phase to accelerate bug-finding.

@@

Figure~\ref{fig:whoop} depicts the \whoop toolchain. Initially, \whoop accepts a Linux driver written in C, together with a Linux environmental model\footnote{Stub header files that model the Linux kernel APIs.}, which is required to ``close'' the driver and allow it to be subsequently over-approximated and analyzed. Both the driver and the environmental model are passed to the translation phase of \whoop (see Section~\ref{whoop:translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}. Next, \whoop performs \emph{static pairwise lockset analysis} to detect all potential data races in the abstract program. The proof sketch behind our static lockset analysis is given in Section~\ref{whoop:proof}. The instrumentation and analysis is described in more detail in Section~\ref{whoop:method}. The summarization that we perform to achieve scalability is presented in Section~\ref{whoop:summarization}. After the static lockset analysis phase, \whoop exploits the race-related information to accelerate bug-finding with \corral (see Section~\ref{whoop:bugfinding})

Finally, we discuss optimizations and assumptions in Section~\ref{whoop:optimizations}, and limitations of our technique in Section~\ref{whoop:limitations}.

@@

The main aim of this phase is to take an input Linux driver written in C, together with an environmental model, and output an over-approximated version of the program in Boogie IVL, a simple imperative language with well-defined verification-focused semantics used as the input to a number of cutting-edge verifiers (e.g. Boogie and \corral). To achieve this translation, \whoop uses three LLVM\footnote{http://llvm.org}-based tools: Chauffeur\footnote{https://github.com/mc-imperial/chauffeur}, Clang\footnote{http://clang.llvm.org} and SMACK\footnote{https://github.com/smackers/smack}~\cite{rakamaric2014smack}.

@@

We now formalize our approach to verifying data race freedom in a driver by statically analyzing its locksets. Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a driver, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this driver. Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all drivers that we have studied so far in the Linux kernel have only a small number of locks (typically one). Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronization bugs that arise from careless fine-grained locking. The actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.

@@

The Engine component of \whoop (see Figure~\ref{fig:whoop}) is responsible for parsing, instrumenting and sequentializing the abstract Boogie program. During the initial parsing, \whoop performs a static analysis (on the Boogie IVL code) to identify all lock identifiers and rewrite them to a unique constant Boogie variable, e.g. the following call to \texttt{mutex\_lock()}:

@@

If \whoop cannot infer a lock it will exit with a warning. A limitation of lockset analysis is that it cannot understand external locking mechanisms (\whoop currently only supports Linux kernel mutexes and spinlocks). It is relatively straightforward to enhance \whoop, though, with new locking primitives. \whoop will also exit with a warning if it detects ``improper'' use of locks (e.g. dynamic lock creation). The reason behind this is twofold: first, it is arguably infeasible to detect such locks using static analysis; and second, because we want to advocate the use of good locking practices when developing drivers for the Linux kernel.

@@

\whoop continues by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialization}, which combines the two arbitrary threads in a single sequential pair. This is achieved by creating a checker function that calls the two instrumented entry points and then performs race checking assertions. This process repeats until all possible pairs of entry points have been sequentialized. To be sound, each time an entry point performs a read access to a shared memory location, \whoop returns a nondeterministic value (using the Boogie keyword \texttt{havoc}). This over-approximates any effects from all the unmodeled threads on the shared state of the pair. Two-thread reduction is not a new idea, it has been used before in GPU kernel verification~\cite{bardsley2014engineering} and in model checking of cache coherence protocols~\cite{mcmillan1999verification}.

@@

Finally the abstract program is send to the Boogie verification engine, which generates verification conditions~\cite{barnett2005weakest} and discharges them to the Z3~\cite{de2008z3} theorem prover. Successful verification implies that the original program is free of data races, while an error denotes a \emph{potential} data race and is reported to the user. To improve usability, \whoop has a built-in error reporter that matches counterexamples to source code. The following is an example of a reported error:

@@

\subsection{Watchdog Summarization}
\label{whoop:summarization}

Early on during the development of \whoop, we realized that scalability would be a serious issue. When we tried to apply \whoop on the r8169 RealTek ethernet driver (8000 lines of C code) with all functions fully inlined, we were running out of memory. One reason was that the C to Boogie translation produced hundreds of thousands of Boogie IVL code, which could not fit in memory. Another reason was that some entry points recursively call helper functions. Inherently, recursion does not work with inlining. To tackle these issues we developed \emph{watchdog summarization}, a novel summarization technique that allows us to analyze Linux drivers in a scalable fashion, but without destroying precision.

Watchdog summarization uses the Houdini~\cite{flanagan2001houdini} invariant inference algorithm (available inside Boogie) to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.

The candidate invariants are based on the global variables that \whoop instruments to perform lockset analysis and race checking (see Section~\ref{whoop:method}). The key idea behind watchdog summarization is that we automatically generate candidate invariants for each memory region in the granularity of watched accesses. This reduces the false positives from reporting errors if only e.g. a field in memory region \texttt{\$M.0} is racing, but the rest of the fields are not. An example of watchdog-based candidate post-conditions is the following:

@@

\whoop is a sound but potentially imprecise static data race analyzer.  While developers do appreciate soundness, it is more important to ensure they are not overwhelmed by countless reports of false bugs.  Hence, to improve precision, we enable an external precise bug-finder for concurrent programs to be plugged into \whoop. The race-related information generated by \whoop can then be used to accelerate the bug-finder.  Currently, we support \corral~\cite{lal2012corral}, a state-of-the-art scalable bug-finder used by Microsoft to analyze Windows device drivers. \corral accepts Boogie IVL and hence it was easy to integrate into our tool flow. Our technique, though, is general and capable of accelerating any arbitrary bug-finder for concurrent programs.

\corral is a symbolic bounded verifier for Boogie IVL that uses the Z3 SMT solver to statically reason about program behaviors. It checks for violations of provided assertions, and reports a precise counterexample if an assertion violation is found. \corral performs bounded exploration of a concurrent program in two steps. First, given a bound on the number of allowed context-switches, the concurrent program is appropriately \emph{sequentialized}, and the generated sequential version preserves reachable states of the original concurrent program~\cite{popl2011-eqr,cav2009-lqr,cavLalR08}. Then, \corral attempts to prove bounded (in terms of the number of loop iterations and recursion depth) sequential reachability of a bug in a goal-directed, lazy fashion to postpone state space explosion when analyzing a large program. It performs two abstractions to achieve this: (i) variable abstraction, where it attempts to identify a minimal set of shared variables that have to be precisely tracked in order to discharge all assertions; and (ii) stratified inlining, where it attempts to inline procedures on-demand as they are required for proving program assertions.

@@

By default, \corral instruments a \texttt{yield} statement (which denotes a
potential nondeterministic context-switch) before each \emph{visible operation}
(i.e., synchronization operations and shared memory accesses) of every thread.
It then leverages sequentialization to explore all possible thread interleavings
up to a pre-defined bound.

Our approach to accelerating \corral is simple and yet effective: we simplify
the sequentialization that \corral performs by instrumenting \texttt{yield}
statements only at shared memory accesses that \whoop reported as potentially
racy. This effectively acts as a partial order reduction that can significantly
reduce the number of interleavings that \corral needs to explore, and thus
increase performance and scalability.

\ZRComment{Could this be cleaned up a bit? In particular, do we need to talk
about entry point pairs, or could we prove it just for any pair?}
We now show that our partial order reduction is sound: it does not miss any more
bugs than what \corral would miss even without taking advantage of the \whoop
race-related information. Let \corral run an $n$ amount of entry point pairs
with yields in all visible operations. Let this configuration be known as $C$.
Each entry point pair is running in a separate \corral process. \corral will
miss all bugs that are triggered by more than 3 bugs running concurrently and
all bugs that are triggered by investigating a larger than $r$ recursion bound
and larger than $k$ context switch bound. Let \corral then run the same $n$
amount of pairs, but this time with all \texttt{yield} statements in non-racy
shared memory accesses (according to \whoop) removed. Let this configuration be
known as $W+C$. Because these shared memory accesses cannot race, performing a
\texttt{yield} will not introduce any new program behaviors. Thus, \corral will
find the same bugs using the configuration $W+C$, as using the configuration
$C$.

@@

\yieldcoarse instruments \texttt{yield} statement in a binary fashion: if \whoop
finds that a pair of entry points cannot race, then it will only instrument
context-switches in synchronization points of this pair; else if \whoop finds
that a pair of entry points can race, then it will instrument context-switches
in all visible operations of this pair.

\yieldmr is a finer-grained instrumentation: \whoop instruments context-switches
in all synchronization points, but only instruments context-switches in memory
regions that can potentially race (regardless if a pair has not been fully
verified as race-free). In our experiments (see Section~\ref{evaluation}),
\yieldmr performs significantly faster than \yieldcoarse.

@@

\subsection{Optimizations and Assumptions}
\label{whoop:optimizations}

A common practice for statically analyzing software is to ``close'' the environment by abstracting away the low level implementation details. Any abstract environmental model, though, can and will ultimately result in decreased precision and thus false positives. This is one of the main limitations of statically analyzing real life software that use complex external APIs. In our case, we developed a simple environmental model for the Linux kernel that consists of stub header files. Because we only focus on lockset-based properties during the race-checking phase of \whoop, we can get away with abstracting a lot of the underlying functionality. Our model, though, can still result in a lot of false positives.

@@

In our experience, one of the most effective precision optimizations is to enrich \whoop with information regarding kernel imposed locks. The Linux kernel is able to serialize a number of entry point calls, which thus can never run concurrently with each other. An an example, a large number of networking and ethernet entry points are mutually serialized with RTNL, a network-specific kernel lock. \whoop will not create pairs for entry points that cannot run concurrently.

Another optimization, is to soundly reduce the checked memory locations. If a memory location is accessed only by one entry point in the pair, then it is safe to not check it. This can potentially decrease the amount of generated candidate invariants and speed up the verification process.

Modeling kernel methods is another opportunity to increase precision, but requires expert knowledge about Linux. For example, the \texttt{register\_netdev()} function registers a network driver with the kernel. Before this function is called, the kernel is not able to invoke any of the driver's entry points. Modeling this functionality can reduce false positives when race checking accesses before the driver's registration.

We also optimize the use of function pointers. LLVM and SMACK do not perform context-sensitive translation, which means that a function that accepts a function pointer as an argument, can arbitrarily call any possible method that corresponds to that function pointer regardless of call site. To increase precision, Chauffeur analyses the AST of the original program and attempts to infer context-sensitive function pointer use information. This information is the used by \whoop to increase precision. To be sound, this optimization is conservative: if no information can be extracted for a function pointer at a call site, then \whoop uses the original over-approximated function.

@@

Environmental modeling is another limitation of \whoop. As aforementioned, enriching \whoop with domain-specific knowledge to increase precision is an ongoing manual effort, but requires expertise in the Linux kernel. We argue that further increasing the precision of our models is orthogonal to the contributions of this paper. Moreover, despite unavoidable imprecisions during the sound static data race analysis, \whoop can use the reported information to significantly speedup precise bug-finders for concurrent programs, as seen in Sections~\ref{whoop:bugfinding} and~\ref{evaluation}.
